State and Parameter Estimation for Spatio-Temporal Neural Fields

### 2010-06-14 ltx/Journal of Neuroengineering/StateAndParameterEstimationNeuralFields_JoNE.tex
paper edits

\begin{abstract}
	This paper presents a framework for creating patient-specific neural field models from electrophysiological data. The Wilson and Cowen or Amari style neural field equations are used to form a parametric model, where the parameters are estimated from data. To illustrate the estimation framework, data is generated using the neural field equations incorporating modelled sensors, so a comparison can be made between estimated and true parameters. To facilitate state and parameter estimation, we introduce a method to reduce the continuum neural field model, using a basis function decomposition, to form a finite dimensional state-space model. Spatial frequency analysis methods are introduced for model selection, by systematically specifying the basis function configuration required to capture the dominant characteristics of the neural field. We use the reduced model to estimate the parameters of the full neural field model \mike{not sure what this last sentence meant?}. The estimation procedure consists of a two-stage iterative algorithm incorporating the unscented Rauch Tung Striebel Smoother for state estimation, and a least squares algorithm for parameter estimation. The results show that it is possible to reconstruct the neural field and estimate intracortical connectivity and synaptic time constant with the proposed framework. The results also illustrate the loss of high spatial frequency information as the cost incurred by the model reduction procedure. This framework provides a link between patient-specific neurophysiological data and theoretical neural fields models. This link may lead to greater understanding of cortical dynamics at the meso/macroscopic scale where diseases such as epilepsy are manifested.   
\end{abstract}

@@

In practice, it is difficult to estimate the bandwidth of the cortex using traditional electrophysiological measurements, stopping us from positioning sensors in accordance to equation~\ref{eq:MinimumSensorDistance}. However, we envisage it may be possible to estimate it using other modalities with higher spatial resolution such as, fMRI, NIRS or other optical imaging techniques~\cite{Issa2000}. Nevertheless, spectral aliasing can still be avoided by a proper choice of the spatial sampling distance given the sensors spectral characteristics. The spatial extent of the sensors results in spectral low-pass action, thus providing spatial anti-aliasing filtering. Such sensors therefore can only sense the field up to a known bandwidth, denoted by $\nu_{cy}$. Aliasing can then be avoided by positioning the sensors according to equation~\ref{eq:MinimumSensorDistance}, with $\nu_{cy}$ replacing $\nu_c$.

@@

\section{Results}\label{ResultsSection} The full neural field model derived in Section~\ref{NeuralModelSection} was used to generate the data for state and parameter estimation. All parameters for the model are given in Table~\ref{tab:Model Parameters}. Figure~\ref{fig:experimental design}(a) shows example data from the sensors of the model, and Figure~\ref{fig:experimental design}(b) shows the power spectral density (PSD) of the simulated time-series. The PSD shows the typical $1/f^2$ characteristics of intracranial EEG.

@@

The sensors were spaced in a regular square 14 $\times$ 14 grid, the sensor width $\sigma^2_m$ was 0.81 which equals $1.5$~mm full width at half maximum, with spacing $\Delta_y$ equal to $1.5$~mm. This arrangement modelled the recording as having some cross talk, which is typical from neurophysiological recordings. %Note, sensor design that is sufficient to guarantee an accurate representation of the field can be achieved using the frequency analysis from Section~\ref{SpectralAnalysisSection} given access to the true field. 

@@

The closest competitor to the integrodifference equation based approach is perhaps the multivariate auto-regressive (AR) model. An advantage of the AR-based methods is that they allow for anisotropic connectivity, which is present in cortico-cortical fibres. It should be noted, however, that the AR and the proposed model-based methods are complementary, where connectivity is estimated at different spatial scales \mike{really? Why can't we use AR models?}. Therefore, we envisage that by combining these methods a more realistic model can be acquired. A recent theoretical study~\cite{Jirsa2009} has demonstrated the importance of this multi-scale structure in generating the characteristic rhythms of the brain.

@@

Parameter estimation may also have significant impact on the treatment of disease. For example, recent theoretical work has demonstrated that seemingly similar electrographic seizures in absence epilepsy may arise from abnormal excitatory or inhibitory mechanisms~\cite{Marten2009}. These very different mechanisms would require very different medications for successful treatment. However, the underlying mechanisms would be hidden from the clinician in the normal clinical setting. Successful parameter estimation has the potential to reveal these hidden mechanisms allowing for improved treatment strategies.

>>>>

\begin{abstract}
	This paper presents a framework for creating patient-specific neural field models from electrophysiological data. The Wilson and Cowen or Amari style neural field equations are used to form a parametric model, where the parameters are estimated from data. To illustrate the estimation framework, data is generated using the neural field equations incorporating modelled sensors, so a comparison can be made between estimated and true parameters. To facilitate state and parameter estimation, we introduce a method to reduce the continuum neural field model, using a basis function decomposition, to form a finite dimensional state-space model. Spatial frequency analysis methods are introduced for model selection, by systematically specifying the basis function configuration required to capture the dominant characteristics of the neural field. The estimation procedure consists of a two-stage iterative algorithm incorporating the unscented Rauch Tung Striebel Smoother for state estimation, and a least squares algorithm for parameter estimation. The results show that it is possible to reconstruct the neural field and estimate intracortical connectivity and synaptic time constant with the proposed framework. The results also illustrate the loss of high spatial frequency information as the cost incurred by the model reduction procedure. This framework provides a link between patient-specific neurophysiological data and theoretical neural fields models. This link may lead to greater understanding of cortical dynamics at the meso/macroscopic scale where diseases such as epilepsy are manifested.   
\end{abstract}

@@

In practice, it is difficult to estimate the bandwidth of the cortex using traditional electrophysiological measurements, possibly preventing us from positioning sensors in accordance to equation~\ref{eq:MinimumSensorDistance}. However, we envisage it may be possible to estimate it using other modalities with higher spatial resolution such as, fMRI, NIRS or other optical imaging techniques~\cite{Issa2000}. Nevertheless, spectral aliasing can still be avoided by a proper choice of the spatial sampling distance given the sensors spectral characteristics. The spatial extent of the sensors results in spectral low-pass action, thus providing spatial anti-aliasing filtering. Such sensors therefore can only sense the field up to a known bandwidth, denoted by $\nu_{cy}$. Aliasing can then be avoided by positioning the sensors according to equation~\ref{eq:MinimumSensorDistance}, with $\nu_{cy}$ replacing $\nu_c$.

@@

\section{Results}\label{ResultsSection} The full neural field model derived in Section~\ref{NeuralModelSection} was used to generate the data for state and parameter estimation. All parameters for the model are given in Table~\ref{tab:Model Parameters}. Note, the parameters for the spatial and temporal aspects of the model can be considered arbitrary when demonstrating the estimation methodology. This is due to the fact that the temporal aspects scale with the sampling rate and the length of the times series and the spatial aspects scale with the size of the neural field and the space step used for numerical simulation. The Mexican hat connectivity kernel is typically comprised of 2 Gaussian basis functions. Figure~\ref{fig:experimental design}(a) shows example data from the sensors of the model, and Figure~\ref{fig:experimental design}(b) shows the power spectral density (PSD) of the simulated time-series. The PSD shows the typical $1/f^2$ characteristics of intracranial EEG.

@@

The sensors were spaced in a regular square 14 $\times$ 14 grid, the sensor width $\sigma^2_m$ was 0.81 which equals $1.5$~mm full width at half maximum, with spacing $\Delta_y$ equal to $1.5$~mm. This arrangement modelled the recording as having some cross talk, which is typical from neurophysiological recordings. 

@@

The closest competitor to the integrodifference equation based approach is perhaps the multivariate auto-regressive (MVAR) model. An advantage of the MVAR-based methods is that they allow for anisotropic connectivity, which is present in cortico-cortical fibres. A recent theoretical study~\cite{Jirsa2009} has demonstrated the importance of this multi-scale structure in generating the characteristic rhythms of the brain. A possible method of creating more precise models with this multi-scale structure maybe to estimate anisotropic cortico-cortical connectivity using a MVAR approach and intracortical connectivity using the methods proposed in this paper.

@@

Parameter estimation may also have significant impact on the treatment of disease. For example, recent theoretical work has demonstrated that seemingly similar electrographic seizures in absence epilepsy may arise from either abnormal excitatory or inhibitory mechanisms~\cite{Marten2009}. These very different mechanisms would require very different medications for successful treatment. However, the underlying mechanisms would be hidden from the clinician in the normal clinical setting. Successful parameter estimation has the potential to reveal these hidden mechanisms allowing for improved treatment strategies.
